---
title: "CC_proj"
author: "Anna Diaz"
date: "2026-03-05"
output: html_document
---

Setup
```{r}

library(tidyverse)
library(scales) # For elegant axis formatting
library(emmeans)
```

Load and Prep Data
```{r}
# 1. Load and prep the data
df <- read_csv("matrix_surrogate_dataset.csv", show_col_types = FALSE) %>%
  # Filter out any infinite or NA values in our target columns
  filter(is.finite(eig_error_relative), is.finite(cond_input)) %>%
  # Create our log-transformed columns and factors for grouping
  rename(dim = size) %>%
  mutate(
    # ggplot treats them as discrete categories
    dim_fct = as.factor(dim),
    scale_fct = as.factor(scale),
    scale_label = paste0("Scale (s) = ", scale)
  )
```

Facet Plot
```{r}
p_multi <- ggplot(df, aes(x = as.factor(log10(cond_input)), y = log10(eig_error_relative), color = dim_fct)) +
  # Jitter and dodge the dots inside each panel
  geom_point(position = position_jitterdodge(jitter.width = 0.2, 
                                             dodge.width = 0.6), 
             alpha = 0.6, size = 2) +
  facet_wrap(~ scale_label) + # Creates the 3 side-by-side panels
  scale_color_viridis_d(option = "viridis") +
  # Format axes to 10^x
  scale_x_discrete(labels = function(x) parse(text=paste0("10^", x))) +
  scale_y_continuous(labels = math_format(10^.x)) +
  labs(
    title = "Error Scaling by Dimension, Condition, and Scale",
    x = expression("Condition Number (" * kappa * ")"),
    y = "Relative Error",
    color = "Dimension (n)"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "right")

ggsave("R_error_multipanel.png", plot = p_multi, width = 12, height = 5, 
       dpi = 300, bg = "white")
print(p_multi)
```

Main Error Plot (All data)
```{r}
p2 <- ggplot(df, aes(x = as.factor(log10(cond_input)), y = log10(eig_error_relative), 
                     color = dim_fct, shape = scale_fct)) +
  # position_jitterdodge adds randomness inside the dodged columns!
  geom_point(position = position_jitterdodge(jitter.width = 0.15, 
                                             dodge.width = 0.6), 
             size = 2.5, alpha = 0.6) +
  scale_color_viridis_d(option = "viridis") +
  # Format axes to 10^x
  scale_x_discrete(labels = function(x) parse(text=paste0("10^", x))) +
  scale_y_continuous(labels = math_format(10^.x)) +
  labs(
    title = "Algorithm Error: Grouped by Dimension and Scale",
    x = expression("Condition Number (" * kappa * ")"),
    y = "Relative Error",
    color = "Dimension (n)", 
    shape = "Scale (s)"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

# Save and show Plot 2
ggsave("R_error_main_plot.png", plot = p2, width = 10, height = 6, dpi = 300,
       bg = "white")
print(p2)
```

Plot the Variance Analysis (Standard Deviation)
```{r}
# Calculate the standard deviation for each group
df_std <- df %>%
  group_by(cond_input, dim, scale, dim_fct, scale_fct) %>%
  summarise(std_log_error = sd(log10(eig_error_relative)), .groups = 'drop')

p1 <- ggplot(df_std, aes(x = as.factor(log10(cond_input)), y = std_log_error, 
                         color = dim_fct, shape = scale_fct)) +
  # position_dodge neatly separates the shapes side-by-side
  geom_point(position = position_dodge(width = 0.5), size = 4, alpha = 0.8) +
  scale_color_viridis_d(option = "viridis") + 
  # Format the X-axis to show 10^x
  scale_x_discrete(labels = function(x) parse(text=paste0("10^", x))) +
  labs(
    title = "Variance Analysis: Grouped by Dimension and Scale",
    x = expression("Condition Number (" * kappa * ")"),
    y = expression("Standard Deviation of " * log[10] * "(Error)"),
    color = "Dimension (n)", 
    shape = "Scale (s)"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

# Save and show Plot 1
ggsave("R_error_std_plot.png", plot = p1, width = 10, height = 6, dpi = 300,
       bg = "white")
print(p1)
```

Mean analysis
```{r}
# Calculate the standard deviation for each group
df_mean <- df %>%
  group_by(cond_input, dim, scale, dim_fct, scale_fct) %>%
  summarise(mean_log_error = mean(log10(eig_error_relative)), .groups = 'drop')

p_mean <- ggplot(df_mean, aes(x = as.factor(log10(cond_input)), y = mean_log_error, 
                              color = dim_fct, shape = scale_fct)) +
  # Add the points on top, dodged to align perfectly
  geom_point(position = position_dodge(width = 0.6), size = 4.5, alpha = 0.9) +
  # Enforce the consistent Viridis color scheme
  scale_color_viridis_d(option = "viridis") + 
  # Format the X-axis to show 10^x
  scale_x_discrete(labels = function(x) parse(text=paste0("10^", x))) +
  labs(
    title = "Average Error: Grouped by Dimension and Scale",
    x = expression("Condition Number (" * kappa * ")"),
    y = expression("Mean " * log[10] * "(Relative Error)"),
    color = "Dimension (n)", 
    shape = "Scale (s)"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "right",
        panel.grid.minor = element_blank())

ggsave("R_error_mean_plot.png", plot = p_mean, width = 10, height = 6, 
       dpi = 300, bg = "white")
print(p_mean)
```

Fitting Linear Regression on Mean Error
```{r}
# We use log10(cond_input) (numeric) to find the slope, and include dim and scale as adjustments
lm_mean <- lm(mean_log_error ~ log10(cond_input) + log10(dim) + log10(scale), 
              data = df_mean)

print("=== Linear Regression Results: MEAN ERROR ===")
summary(lm_mean)

# Linear Regression on the Standard Deviation (Variance)
lm_std <- lm(std_log_error ~ log10(cond_input) + dim_fct + scale_fct, 
             data = df_std)

print("=== Linear Regression Results: STANDARD DEVIATION ===")
summary(lm_std)
```

Fitting an ANOVA model for average error (linear regression averages the trails)
Things to consider don't treat scale and dimension as categorical factors.
Treat log condition number, scale, and dim all as continuous as they are numeric. 
```{r}
#Writing the math directly into the formula using the original numeric columns!
continuous_model <- lm(log10(eig_error_relative) ~ log10(cond_input) 
                       * log10(dim)
                       * log10(scale), data = df)

print("=== Running Stepwise Selection (Targeting lowest AIC) ===")
final_model <- step(continuous_model, direction = "both", trace = 1)

print("=== Final Optimized Model ===")
summary(final_model)

```
Can check residual plots 
```{r}
plot(final_model)
```
Confirms our data is non-linear because of the spike at the end as previously
mentioned. We may want to consider looking at the data without the other group 

Fit the final model to the plot of relative error
```{r}
# Calculate the continuous model predictions (assuming final_model is already run)
df$predicted_mean <- predict(final_model, newdata = df)

# Plot the Continuous Mean Model 
p_mean_fit <- ggplot(df, aes(x = as.factor(log10(cond_input)))) +
  
  # THE DOTS: Colored by Dimension, Shaped by Scale, jittered/dodged for visibility
  geom_point(aes(y = log10(eig_error_relative), color = as.factor(dim), shape = as.factor(scale)), 
             position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.6), 
             size = 2.5, alpha = 0.5) +
  
  # THE MODEL LINES: Forced to group ONLY by Scale, drawn as crisp black lines
  # We use linetype to distinguish the three scales. 
  geom_line(aes(y = predicted_mean, group = as.factor(scale), linetype = as.factor(scale)), 
            linewidth = 1.2, color = "black") +
  
  scale_color_viridis_d(option = "viridis") + 
  scale_x_discrete(labels = function(x) parse(text=paste0("10^", x))) +
  labs(
    title = "Continuous Linear Model: Mean Relative Error",
    x = expression("Condition Number (" * kappa * ")"),
    y = expression("log[10] * (Relative Error)"),
    color = "Dimension (n)", 
    shape = "Scale (s)",
    linetype = "Model Scale (s)" # Links the lines in the legend
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "right")

# Save and show the plot
ggsave("R_continuous_model_fit.png", plot = p_mean_fit, width = 10, height = 6, 
       dpi = 300, bg = "white")
print(p_mean_fit)

```

The model is able to bend itself when needed. As previously mentioned the 
lines are slanted becuase of the big leap in the condition number 10^12, where
it is not constant from the rest of the data. 


DONT TRY UNTIL WE HAVE CHECKED THE MODEL THOROUGHLY 
Look into contrasts: to answer question about "at specific scale, 
does this one dimension have a significantly different error slope than this 
other dimension?" function emtrends (because everything is numerical) 
```{r}
#We want to see the TREND (slope) of log condition number
# at specific continuous values of scale 0.01, 1, and 100

final_trends <- emtrends(final_model, 
                         ~ scale, 
                         var = "log10(cond_input)", 
                         at = list(scale = c(0.01, 1, 100)))

print("=== Linear Slopes of Condition Number by Scale ===")
print(final_trends)

# Test if these slopes are statistically different from 0 (a flat line)
test(final_trends)

# Pairwise comparisons of the slopes 
contrast(final_trends)

```
The contrasts show that the slopes are statistically different from each other. 
But I believe it may be due to that one group because as we saw before 
everything else looks constant but that the stuff in condition number 10^12. 


Stepwise Selection for Variance Analysis
we saw previously that the 3 way interactions were not significant
so in choosing a model for the variance analysis, we can use stepwise to proceed. 
```{r}
full_std_model <- lm(std_log_error ~ log10(cond_input) * log10(dim) 
                     * log10(scale), data = df_std)

# run the stepwise 
print("=== Running Stepwise Selection (Targeting lowest AIC) ===")
best_std_model <- step(full_std_model, direction = "both", trace = 1)

print("=== Final Optimized Variance Model ===")
summary(best_std_model)

# generate predictions using the selected BEST model
df_std$predicted_std <- predict(best_std_model, newdata = df_std)

```

Residuals for Variance model
```{r}
plot(best_std_model)
```
Residuals don't violate any assumptions.

Final Variance model Plot
```{r}
# Plot the Continuous Variance Model (Lines grouped by Scale)
p_std_fit <- ggplot(df_std, aes(x = as.factor(log10_cond))) +
  
  # THE DOTS: Colored by Dimension, Shaped by Scale, jittered/dodged for visibility
  geom_point(aes(y = std_log_error, color = dim_fct, shape = scale_fct), 
             position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.6), 
             size = 4, alpha = 0.5) +
  
  # THE MODEL LINES: Forced to group ONLY by Scale, drawn as crisp black lines
  # We use linetype to distinguish the three scales.
  geom_line(aes(y = predicted_std, group = scale_fct, linetype = scale_fct), 
            linewidth = 1.2, color = "black") +
  
  scale_color_viridis_d(option = "viridis") + 
  scale_x_discrete(labels = function(x) parse(text=paste0("10^", x))) +
  labs(
    title = "Continuous Linear Model: Variance",
    x = expression("Condition Number (" * kappa * ")"),
    y = expression("Standard Deviation of " * log[10] * "(Relative Error)"),
    color = "Dimension (n)", 
    shape = "Scale (s)",
    linetype = "Model Scale (s)" # Links the lines in the legend
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "right")

# Save and show the plot
ggsave("R_continuous_var_model_fit.png", plot = p_std_fit, width = 10, height = 6, 
       dpi = 300, bg = "white")
print(p_std_fit)

```

ALTERNATIVE METHOD: Averaging the Means (Marginal MEAN)
Calculate the Grand Mean and show why it wouldn't be a good choice to do.
Averaging the means would flatten the data and ignore the specific conditions. 
```{r}
# 1. Calculate the Grand Mean 
# We group by condition number and dimension, completely ignoring the scales.
df_grand_mean <- df_mean %>%
  group_by(cond_input, dim_fct) %>%
  summarise(grand_mean_error = mean(mean_log_error), .groups = 'drop')

# 2. Plot the Flawed Grand Mean
p_grand <- ggplot(df_grand_mean, aes(x = as.factor(log10(cond_input)), 
                                     y = grand_mean_error, 
                                     color = dim_fct)) +
  # Add points and lines to show the false trend
  geom_point(position = position_dodge(width = 0.5), size = 4.5, alpha = 0.9) +
  geom_line(aes(group = dim_fct), position = position_dodge(width = 0.5), 
            linewidth = 1) +
  scale_color_viridis_d(option = "viridis") + 
  scale_x_discrete(labels = function(x) parse(text=paste0("10^", x))) +
  labs(
    title = "Grand Mean Error (Averaged Across Scales)",
    x = expression("Condition Number (" * kappa * ")"),
    y = expression("Grand Mean " * log[10] * "(Relative Error)"),
    color = "Dimension (n)"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "right")

# Save and show the plot
ggsave("R_flawed_grand_mean_plot.png", plot = p_grand, width = 10, height = 6, 
       dpi = 300, bg = "white")
print(p_grand)
```

Now we are not seeing certain details that we saw in previous graphs. Now the 
graph looks like it's entirely constant until 10^12 but we saw in one graph
that for certain scales 10^12 was still constant until a certain scale. 
Plotting the grand mean revealed that averaging across the scales diluted the
error metrics. This is masking the critcial interaction of each scale. Therefore,
to preserve the behavior of the solver, all averages and linear models were
calculated strictly at the replicate level within their specific scale and
dimension. 


Create model with additional variable indicator (10^12)
Model with variable indicator (MEAN analysis)
```{r}
# Creating a TRUE/FALSE column, treating it as 1 or 0 (dummy variable)
df$is_failure <- ifelse(round(log10(df$cond_input)) == 12, 1, 0)

# The scale behaves differently in condition number 10^12, thus we will
# include the interaction in the model

indicator_model <- lm(log10(eig_error_relative) ~ log10(dim) + 
                        log10(cond_input) * log10(scale) + is_failure + 
                        is_failure:log10(scale), data = df)

print("=== Running Stepwise Selection on Indicator Model ===")
best_indicator_model <- step(indicator_model, direction = "both", 
                             trace = 1)

print("=== Final Optimized Indicator Model ===")
summary(best_indicator_model)
```

Now graph the indicator model 
```{r}
# calculate the predictions 
df$predicted_log_error <- predict(indicator_model, df)

p_indicator <- ggplot(df, aes(x = as.factor(log10(cond_input)))) +
  
  # THE DOTS: Colored by Dimension, Shaped by Scale, dodged for visibility
  geom_point(aes(y = log10(eig_error_relative), color = as.factor(dim), 
                 shape = as.factor(scale)), 
             position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.6), 
             size = 2.5, alpha = 0.5) +
  
  # THE MODEL LINES: One crisp black line per scale, mapping the structural break!
  geom_line(aes(y = predicted_log_error, group = as.factor(scale), 
                linetype = as.factor(scale)), 
            linewidth = 1.2, color = "black") +
  
  # Colors and X-axis formatting
  scale_color_viridis_d(option = "viridis") + 
  scale_x_discrete(labels = function(x) parse(text=paste0("10^", x))) +
  
  # Clean up labels
  labs(
    title = "Structural Break Model: Mean Relative Error",
    x = expression("Condition Number (" * kappa * ")"),
    y = expression("log[10] * (Relative Error)"),
    color = "Dimension (n)", 
    shape = "Scale (s)",
    linetype = "Model Scale (s)" # Matches the legend for the lines to the shapes
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "right")

ggsave("Indicator_Model_Plot.png", plot = p_indicator, width = 10, height = 5, 
       bg = "white")
print(p_indicator)
```

Indicator model for Variance Analysis
```{r}
df_std <- df_std %>%
  mutate(
    log10_cond = log10(cond_input),
    log10_scale = log10(scale),
    log10_dim = log10(dim),
    is_failure = ifelse(round(log10_cond) == 12, 1, 0)
  )
 
sd_indicator_model <- lm(std_log_error ~ log10_dim + log10_cond * log10_scale + 
                                         is_failure + is_failure:log10_scale, 
                         data = df_std)

print("=== Structural Break Model: Standard Deviation ===")
best_struct <- step(sd_indicator_model, direction = "both", trace = 1)
print("=== Final Optimized Indicator Model ===")
summary(best_struct)
```

Now the graph for the Variance Analysis (we don't need this one)
```{r}
# 1. Calculate the predictions (you already ran this, but just to be safe)
df_std$predicted_std <- predict(best_struct, newdata = df_std)

# 2. Plot the Variance Indicator Model (Grouped by Dimension, not Scale)
p_var_break <- ggplot(df_std, aes(x = as.factor(log10_cond))) +
  
  # THE DOTS: Colored by Dimension, Shaped by Scale
  geom_point(aes(y = std_log_error, color = dim_fct, shape = scale_fct), 
             position = position_dodge(width = 0.5), 
             size = 4, alpha = 0.4) +
  
  # THE MODEL LINES: Split and colored by Dimension!
  # (Because dimension survived the stepwise cut, we draw 4 lines to match the 4 sizes)
  geom_line(aes(y = predicted_std, group = dim_fct, color = dim_fct), 
            linewidth = 1.5) +
  
  scale_color_viridis_d(option = "viridis") + 
  scale_x_discrete(labels = function(x) parse(text=paste0("10^", x))) +
  labs(
    title = "Structural Break Model: Variance (Standard Deviation)",
    x = expression("Condition Number (" * kappa * ")"),
    y = expression("Standard Deviation of " * log[10] * "(Relative Error)"),
    color = "Dimension (n)", 
    shape = "Scale (s)"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "right")

# Save and show the plot
ggsave("R_var_indicator_model_fit.png", plot = p_var_break, width = 10, height = 6, 
       dpi = 300, bg = "white")
print(p_var_break)
```

